\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Quaternion-Based IR for Quantum Circuit Optimization}

\author{\IEEEauthorblockN{John Van Geem}
\IEEEauthorblockA{\textit{CEO, RQM Technologies}\\
Arlington, VA, USA\\
0009-0002-4003-8452}
}

\maketitle

\begin{abstract}
Quantum circuit optimizers often operate on named gate syntax, so semantically
identical single-qubit operations appear as different instructions across frameworks
and backends. This paper presents a quaternion-based intermediate representation (IR)
for quantum circuit optimization that normalizes all single-qubit gates into one common
form before optimization passes are applied. Because single-qubit operations form the
group SU(2), unit quaternions provide an algebraically exact and compact encoding. The
representation enables composition, simplification, and backend lowering to be performed
on gate semantics rather than on syntactic aliases. We explain how this change improves
compiler clarity, supports backend-agnostic optimization, and provides a cleaner
foundation for gate fusion and redundancy elimination. The result is a more principled
optimization boundary than standard named gate-operation pipelines.
\end{abstract}

\begin{IEEEkeywords}
quantum circuit optimization, intermediate representation, SU(2), quaternions,
compiler design, backend-agnostic compilation, gate fusion
\end{IEEEkeywords}

%% ============================================================
\section{Introduction}
%% ============================================================

Quantum compiler pipelines must handle a growing diversity of gate sets, native bases,
and backend conventions. As hardware providers multiply and high-level frameworks
proliferate, a compilation layer that normalizes quantum programs into a stable internal
form becomes essential for correctness, portability, and long-term maintainability. The
challenge is not merely that different backends use different native gates; it is that
the same physical operation is routinely expressed in distinct syntactic forms, and
current compilers are forced to reason about those forms rather than about the underlying
physics.

The central problem is that current compilers reason about single-qubit operations at the
level of \textit{named gate syntax}. A rotation by angle $\theta$ about the $Z$-axis can
appear as \texttt{rz}$(\theta)$, \texttt{p}$(\theta)$, \texttt{u1}$(\theta)$, or a phase
gate --- all semantically identical but syntactically distinct. A compiler that operates
over gate names must maintain a growing equivalence table: adding a new gate alias requires
updating the table, and supporting a new backend requires translating each named gate
individually. Complexity scales with the number of gate aliases, not with the mathematical
structure of the operations.

This syntax-centeredness has a subtle but important consequence for how optimization
quality is measured. A syntax-level optimizer can only identify equivalences that its
alias table explicitly anticipates. If two circuits are written using different gate
names for the same rotation, the optimizer may fail to recognize the cancellation
opportunity. The quality of optimization therefore depends on the completeness of
a manually curated enumeration, not on any intrinsic property of the computation. This
is a fragile foundation for a compiler that must remain correct across rapidly evolving
hardware generations and framework ecosystems.

A compiler IR should be evaluated not only by implementation convenience but also by
\textit{semantic stability}: how well the internal representation isolates the optimizer
from surface-level syntactic variation. A representation is semantically stable if two
circuits that compute the same physical operation are guaranteed to have the same IR form,
regardless of which gate names or parameter conventions were used to write them. Semantic
stability is the property that makes optimization passes reliable across frameworks and
backends. Without it, optimizers must be continuously extended as new syntactic variants
appear, and correctness guarantees become difficult to state precisely.

This paper argues that the correct normalization boundary for single-qubit optimization is
at the level of \textit{SU(2) geometry}. Single-qubit operations form the Lie group SU(2),
which is isomorphic to the unit 3-sphere $S^3$ and admits a compact, algebraically exact
representation as unit quaternions. A compiler that normalizes all single-qubit gates into a
quaternionic IR can reason about equivalence, composition, and simplification using the
mathematics of the operation rather than a register of names. This normalization is not an
approximation: the isomorphism between SU(2) and unit quaternions is exact, and quaternion
multiplication faithfully reproduces the composition law of single-qubit unitary matrices.

The central contribution of this paper is a quaternion-based IR, which we call \textbf{u1q},
serving as a \textit{semantic normalization boundary} for single-qubit optimization. By
establishing this boundary before any optimization pass runs, the compiler guarantees that
all single-qubit gates --- from any framework, in any parameterization --- arrive at the
optimizer in an identical canonical form. We describe the mathematical structure of this
representation, define its invariants, and argue concretely why it enables cleaner gate
fusion, identity elimination, and backend-agnostic lowering than syntax-level pipelines
support. We also describe a working implementation that validates the practicality of the
approach.

%% ============================================================
\section{Why Standard Gate-Based Optimization Is Limited}
%% ============================================================

\subsection{The Gate-Alias Problem}

Modern quantum frameworks each define their own named gate vocabularies. Qiskit, Amazon
Braket, and PennyLane express the same physical operation --- a phase shift, a
parameterized rotation, a Hadamard --- using different instruction names and parameter
conventions. Even within a single framework, gate names alias one another: Qiskit's
\texttt{rz}, \texttt{p}, and \texttt{u1} all implement a $Z$-axis rotation with different
historical names, different default parameter interpretations in older API versions, and
slightly different handling in transpiler passes.

A compiler that optimizes over gate names must explicitly enumerate which names are
equivalent before any optimization pass can reason about them. This is not a one-time cost:
each new framework or backend extends the alias table, and each new optimization pass
must be aware of the full table. The result is a compiler whose complexity grows with
syntactic diversity rather than with the mathematical structure of the operations being
optimized. For any given alias pair, the developer must decide whether the equivalence is
exact or context-dependent, whether it holds under all parameter values, and whether it
propagates through subsequent gates. These questions have straightforward answers in the
algebra of SU(2); they are much harder to answer systematically in a name-based registry.

\subsection{Backend Naming Divergence}

The naming problem is compounded by backend divergence. Hardware providers impose their
own native gate vocabularies, and these vocabularies change between device generations.
A compiler targeting IBM hardware works with gates like \texttt{cx}, \texttt{rz}, and
\texttt{sx}; a compiler targeting IonQ or Quantinuum works with a different set of native
primitives. A circuit optimized in one basis must be retranslated into another, and the
translation cannot reuse optimization results from the first basis without re-deriving
the equivalences.

This creates a maintenance structure in which backend-specific optimization logic is
duplicated rather than shared. Each new backend requires not just a new translation table
but also a new set of rewrite rules, because the semantic content of the optimization
must be re-expressed in each new syntactic vocabulary. The problem is structural: as long
as the optimization boundary is at the level of gate names, optimization logic is
inherently backend-specific.

\subsection{The Maintenance Burden of Syntax-Level Rewrite Systems}

Pattern-matching rewrite rules over gate names face a combinatorial problem. The number
of rewrite patterns needed to cover all alias combinations grows quadratically with the
number of gate names, and each backend introduces its own naming conventions. Compilers
that rely on this approach require continuous maintenance as gate sets evolve. A rewrite
system that was complete for a given framework at one point in time becomes incomplete
the moment a new gate alias is introduced, a parameter convention is updated, or a new
backend is added to the supported target list.

Beyond scalability, the engineering burden is asymmetric: every new gate name
requires reasoning about how it interacts with every existing rewrite rule. This
reasoning is non-trivial because gate equivalences under composition are not always
obvious from gate names alone. A human reviewer inspecting a rewrite system cannot easily
verify its completeness without independently knowing the full algebra of the gate set.

\subsection{Syntax-Level Equivalence Is Not Semantic Equivalence}

It is important to distinguish two notions of equivalence. \textit{Syntactic equivalence}
holds between two circuit representations that use the same gate names and parameters.
\textit{Semantic equivalence} holds between two circuit representations that compute the
same physical unitary, regardless of how they are written. A compiler operating at the
syntax level can only detect syntactic equivalences that its rewrite rules explicitly
anticipate. Two circuits that are semantically equivalent --- because they represent the
same sequence of physical rotations --- may be syntactically distinct in a way the
compiler has no rule to recognize.

This gap between syntactic and semantic equivalence is not merely theoretical. It arises
in practice whenever the same rotation is expressed using different gate types (e.g.,
\texttt{rz} versus \texttt{p}), different parameter conventions (e.g., full-angle versus
half-angle), or different framework vocabularies. An optimizer that misses these cases
will produce sub-optimal output that is correct but not as short as it could be. An
optimizer that incorrectly classifies a syntactic form as a known alias can produce
output that is incorrect. Both failure modes are consequences of the same structural
decision: locating the optimization boundary at the syntactic rather than the semantic
level.

\subsection{Brittleness of Syntax-Oriented Optimization}

The foregoing observations converge on a single architectural critique: syntax-oriented
quantum compilers produce brittle optimization logic. By ``brittle'' we mean that the
correctness and completeness of the optimizer depend on properties of the input
representation --- specifically, on which gate names were used and whether they are
covered by the alias table --- rather than on the mathematical content of the circuit.
Two circuits that are physically identical may be optimized differently, or not at all,
simply because they were written using different names. A compiler with this property
cannot offer uniform guarantees across frameworks, and it requires non-trivial engineering
effort to extend to new backends or gate sets.

The natural remedy is to move the optimization boundary below the syntactic level --- to
a representation that encodes gate semantics directly. The quaternion IR proposed in this
paper is such a representation. It eliminates gate aliases by construction, makes
semantic equivalence structurally visible, and supports a uniform set of optimization
rules that apply to all single-qubit gates simultaneously.

%% ============================================================
\section{Quaternion-Based Intermediate Representation}
%% ============================================================

\subsection{Mathematical Foundation}

A single-qubit gate is a $2{\times}2$ complex unitary matrix with determinant 1 --- an
element of the special unitary group SU(2). The group SU(2) has a well-known geometric
interpretation: it is isomorphic to the unit 3-sphere $S^3$ in four-dimensional Euclidean
space, meaning that every single-qubit rotation corresponds to a unique point on $S^3$
(up to a global sign, discussed below). This geometric structure is not just a
mathematical curiosity; it means that the space of single-qubit rotations is compact,
connected, and admits exact arithmetic using four real numbers rather than four complex
numbers (eight real parameters constrained to a lower-dimensional manifold).

A general rotation by angle $\theta$ about unit axis
$\hat{n} = (n_x, n_y, n_z)$ corresponds to the SU(2) element:
\begin{equation}
U(\hat{n}, \theta) = \cos\!\tfrac{\theta}{2}\,I - i\sin\!\tfrac{\theta}{2}\,(\hat{n} \cdot \boldsymbol{\sigma}),
\label{eq:su2}
\end{equation}
where $\boldsymbol{\sigma} = (\sigma_x, \sigma_y, \sigma_z)$ are the Pauli matrices.
The factor of $\tfrac{1}{2}$ in the angle arguments reflects the spinorial nature of
SU(2): a physical rotation by $2\pi$ returns a qubit state to itself up to a global phase,
while the SU(2) matrix requires a full $4\pi$ rotation to return to the identity. This
is the double-cover structure of SU(2) over the rotation group SO(3), and it is important
for the canonicalization conventions described below.

Every unit quaternion
\begin{equation}
q = w + xi + yj + zk, \quad w^2 + x^2 + y^2 + z^2 = 1
\label{eq:quat}
\end{equation}
corresponds exactly to one SU(2) element via the standard identification:
\begin{equation}
U(q) =
\begin{bmatrix}
 w + iz  &  y + ix \\
-y + ix  &  w - iz
\end{bmatrix}.
\label{eq:su2matrix}
\end{equation}
This correspondence is an algebraic isomorphism, not a coordinate approximation.
Every arithmetic operation on quaternions --- multiplication, conjugation, norm
computation --- has a direct counterpart in SU(2) matrix algebra. Quaternion
multiplication corresponds exactly to matrix multiplication, and the conjugate $q^*$
corresponds exactly to the matrix adjoint $U^\dagger$. A compiler that works with
quaternions is therefore working with the exact mathematical structure of the gates,
using a representation that is both more compact and more computationally convenient than
the full matrix form.

The axis-angle decomposition of a quaternion takes the form:
\begin{equation}
q = \cos\!\tfrac{\theta}{2} + \hat{u}\sin\!\tfrac{\theta}{2},
\label{eq:axisangle}
\end{equation}
where $\hat{u} = u_x i + u_y j + u_z k$ is a unit pure quaternion encoding the rotation
axis. Given any quaternion $q = (w, x, y, z)$, the rotation angle is $\theta = 2\arccos(w)$
and the rotation axis is $\hat{n} = (x, y, z) / \sin(\theta/2)$ when $\theta \neq 0$.
This decomposition provides a direct path from quaternion form to axis-angle parameters
and back, making backend decomposition (e.g., ZYZ decomposition for IBM hardware) a
well-defined closed-form operation.

\subsection{Why Quaternions Are a Natural Compiler Representation}

Three properties of unit quaternions make them particularly well suited as a compiler IR
for single-qubit gates.

First, \textit{closure under composition}: the product of two unit quaternions is again a
unit quaternion. This means that gate fusion --- the composition of two single-qubit
operations --- never leaves the representation space. There is no intermediate form, no
normalization step between operations, and no risk of accumulating a non-canonical
intermediate. Sequential single-qubit gates can be fused into a single quaternion by a
sequence of multiplications, and the result is immediately in the canonical representation.

Second, \textit{compactness}: a unit quaternion requires exactly four real numbers subject
to one quadratic constraint, for an effective dimension of three --- matching the three
degrees of freedom of SU(2). The full $2\times2$ complex matrix representation uses eight
real numbers (or four complex) subject to four constraints. The quaternion form is not just
more convenient computationally; it is the minimal-redundancy representation of the gate's
physical content.

Third, \textit{algebraic exactness}: the quaternion representation is not a parameterization
that introduces approximation error. The isomorphism between unit quaternions and SU(2) is
exact, and quaternion arithmetic is exact subject only to floating-point precision. Gate
composition using quaternion multiplication preserves the same precision as matrix
multiplication and does not accumulate additional approximation. For compiler verification
purposes --- confirming that an optimized circuit computes the same unitary as the
original --- the quaternion form supports exact norm checks and equality tests with
well-defined numerical tolerances.

\subsection{IR Definition and Invariants}

The quaternion IR, which we call \textbf{u1q}, encodes every single-qubit gate as a unit
quaternion $(w, x, y, z)$ satisfying the constraints in Table~\ref{tab:invariants}.

\begin{table}[htbp]
\caption{u1q Canonicalization Invariants}
\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Constraint} & \textbf{Value} \\
\midrule
Unit norm      & $|w^2 + x^2 + y^2 + z^2 - 1| < 10^{-9}$ \\
Non-degenerate & Not all components identically zero \\
Canonical sign & $w \geq 0$ (geodesic shortest arc) \\
\bottomrule
\end{tabular}
\label{tab:invariants}
\end{center}
\end{table}

The unit-norm constraint enforces that the quaternion lies on $S^3$, corresponding to a
valid SU(2) element. The non-degeneracy constraint rules out the zero quaternion, which
has no geometric interpretation. The canonical-sign constraint --- $w \geq 0$ --- resolves
the SU(2) double-cover ambiguity: both $q$ and $-q$ represent the same physical rotation,
so the IR must select exactly one representative. The convention $w \geq 0$ places the
representative in the upper hemisphere of $S^3$, corresponding to the rotation arc of
length at most $\pi$ (the physically shorter rotation), and matches standard conventions
in quaternion interpolation literature.

Together, these invariants define a unique canonical form for every single-qubit unitary.
Any two gates that compute the same physical rotation --- regardless of their original
names, frameworks, or parameter conventions --- will produce the same u1q quaternion after
canonicalization. This is the semantic normalization property that makes the IR valuable
as an optimization boundary.

\subsection{Gate-to-Quaternion Mapping}

Table~\ref{tab:gatemapping} gives the canonical quaternion coordinates for a representative
set of standard single-qubit gates, derived from \eqref{eq:axisangle} with the $w \geq 0$
sign convention. The table illustrates how diverse gate types --- discrete Clifford gates,
parameterized rotations, and multi-angle gates --- all map into the same four-component form.

\begin{table*}[htbp]
\renewcommand{\arraystretch}{1.35}
\caption{Canonical Gate-to-Quaternion Mapping}
\begin{center}
\begin{tabular}{lccccl}
\toprule
\textbf{Gate} & $w$ & $x$ & $y$ & $z$ & \textbf{Notes} \\
\midrule
Identity          & $1$ & $0$ & $0$ & $0$
  & No rotation; identity in SU(2) \\[2pt]
X                 & $0$ & $1$ & $0$ & $0$
  & $\pi$-rotation about $x$-axis \\[2pt]
Y                 & $0$ & $0$ & $1$ & $0$
  & $\pi$-rotation about $y$-axis \\[2pt]
Z                 & $0$ & $0$ & $0$ & $1$
  & $\pi$-rotation about $z$-axis \\[2pt]
H                 & $\dfrac{1}{\sqrt{2}}$ & $\dfrac{1}{\sqrt{2}}$ & $0$ & $0$
  & $\pi$-rotation about $(x{+}z)/\sqrt{2}$ \\[8pt]
S                 & $\dfrac{1}{\sqrt{2}}$ & $0$ & $0$ & $\dfrac{1}{\sqrt{2}}$
  & $R_z(\pi/2)$; S gate \\[8pt]
T                 & $\cos\!\tfrac{\pi}{8}$ & $0$ & $0$ & $\sin\!\tfrac{\pi}{8}$
  & $R_z(\pi/4)$; T gate \\[6pt]
$R_x(\theta)$     & $\cos\!\tfrac{\theta}{2}$ & $\sin\!\tfrac{\theta}{2}$ & $0$ & $0$
  & Rotation by $\theta$ about $x$-axis \\[4pt]
$R_y(\theta)$     & $\cos\!\tfrac{\theta}{2}$ & $0$ & $\sin\!\tfrac{\theta}{2}$ & $0$
  & Rotation by $\theta$ about $y$-axis \\[4pt]
$R_z(\theta)$     & $\cos\!\tfrac{\theta}{2}$ & $0$ & $0$ & $\sin\!\tfrac{\theta}{2}$
  & Rotation by $\theta$ about $z$-axis; same as \texttt{rz}, \texttt{p}, \texttt{u1} \\[4pt]
$U(\theta,\phi,\lambda)$ & \multicolumn{4}{c}{derived via ZYZ decomposition}
  & General single-qubit gate \\[2pt]
\bottomrule
\end{tabular}
\label{tab:gatemapping}
\end{center}
\end{table*}

The $R_z(\theta)$ row illustrates the alias-collapse property directly: \texttt{rz},
\texttt{p}, and \texttt{u1} all map to the same quaternion $(\cos(\theta/2), 0, 0, \sin(\theta/2))$,
making their equivalence structurally visible at the IR level. The general unitary
$U(\theta, \phi, \lambda)$ is handled by first computing the corresponding SU(2) matrix
via the ZYZ Euler decomposition and then extracting the quaternion components from
\eqref{eq:su2matrix}. This covers the complete set of single-qubit gates used across
current frameworks.

The canonicalization pass \texttt{to\_u1q\_pass} traverses the instruction list and maps
every named single-qubit gate to its u1q equivalent. Two-qubit gates (CNOT, CZ, SWAP)
are left unchanged and serve as segment boundaries: the single-qubit optimization domain
is a maximal sequence of single-qubit operations on a given qubit between consecutive
two-qubit gate applications. After this pass, the IR contains only u1q nodes and explicit
two-qubit operations, and the compiler can proceed with optimization at the semantic level.

%% ============================================================
\section{Why Quaternion-Based IR Is an Upgrade}
%% ============================================================

This section is the argumentative center of the paper. Each advantage described here
is structural, following from the algebraic isomorphism between unit quaternions and
SU(2), not from heuristic engineering decisions. Taken together, they support the
claim that the quaternion IR is not simply an alternative encoding but a better
optimization boundary.

\subsection{Canonical Semantic Normalization}

After \texttt{to\_u1q\_pass}, every single-qubit gate --- regardless of its original name,
framework, or parameter convention --- has been mapped to a single canonical form. Two
gates that are physically identical produce the same quaternion. This is a stronger
guarantee than any alias table can provide, because it is grounded in the mathematical
structure of SU(2) rather than in a manually maintained enumeration.

The significance of this guarantee is easiest to see at a system level. A compiler
equipped with an alias table can guarantee that its \textit{known} aliases are handled
consistently; it cannot guarantee that \textit{all} semantically equivalent inputs are
treated identically. The u1q canonicalization makes no reference to gate names at all.
Two inputs are treated identically if and only if they compute the same SU(2) element,
which is an intrinsic mathematical property of the gates rather than a property of
the names assigned to them in a given framework.

\subsection{Collapse of Gate Aliases}

Name-based compilers must explicitly handle each alias pair: (\texttt{rz}, \texttt{p}),
(\texttt{p}, \texttt{u1}), (\texttt{u1}, \texttt{phaseshift}), and so on. The quaternion
IR eliminates this entirely. Once all gates are in u1q form, there are no aliases ---
only quaternions. Every optimization pass that operates over u1q automatically covers the
complete space of single-qubit operations, without any awareness of the gate names that
appeared in the input.

This has a concrete engineering consequence. When a new framework introduces a new gate
alias --- say, a \texttt{phase\_shift} gate that is physically identical to \texttt{rz} ---
a name-based compiler must be extended to recognize the new alias in every optimization
pass that might benefit from it. A quaternion IR compiler requires only one update: the
gate resolution step that maps the new gate name to its quaternion coordinates. All
downstream optimization passes automatically handle the new gate correctly without any
modification.

\subsection{Backend-Agnostic Optimization Boundary}

Fig.~\ref{fig:pipeline} illustrates the structural difference between the two compilation
strategies. In a standard named-gate pipeline, optimization logic must be re-applied or
re-expressed for each backend because each backend uses different gate names. Optimization
that was applied in a framework-native gate set may need to be partially re-done after
translation to a hardware-native gate set, because the gate names and rewrite rules have
changed.

In the quaternion IR pipeline, all single-qubit optimization occurs at the u1q level
before any backend lowering. Each backend adapter receives a semantically optimized u1q
IR and translates it once, using a fixed decomposition (such as ZYZ for IBM hardware or
axis-angle for trapped-ion systems), without re-deriving gate semantics or re-running
optimization. The separation between the optimization boundary and the backend translation
step is clean and enforced by the IR design: optimization produces u1q, and backend
adapters consume u1q. Neither stage has any knowledge of the other's naming conventions.

This separation also makes the compiler more predictable: the degree of optimization
achieved is a function of the circuit's semantic content, not of the backend being
targeted. A circuit that can be reduced by quaternion fusion will be reduced regardless
of whether it is ultimately lowered to IBM, IonQ, or any other backend.

\subsection{Cleaner Fusion and Inversion Rules}

Gate fusion in a named-gate compiler requires a growing library of type-specific rewrite
rules. For example: $R_z(\alpha) \cdot R_z(\beta) \to R_z(\alpha{+}\beta)$,
$H \cdot H \to I$, $S \cdot S^\dagger \to I$, $R_x(\alpha) \cdot R_x(\beta) \to R_x(\alpha{+}\beta)$,
and so on. Each of these rules must be stated separately, and rules that mix gate types
(e.g., composing $R_x$ and $R_y$) are more complex still.

In the quaternion IR, all of these rules collapse into a single operation. Gate fusion
is quaternion multiplication:
\begin{equation}
q_{\text{fused}} = q_2 \cdot q_1,
\label{eq:fusion}
\end{equation}
where $\cdot$ is quaternion multiplication, applied left-to-right in circuit order.
This single rule covers the complete space of single-qubit gate sequences, including
composites of gates on different axes and arbitrary parameterized rotations. There is no
case analysis over gate types. The fusion result is immediately in canonical u1q form,
ready for subsequent passes.

Gate inversion is equally direct. For any unit quaternion, $q^{-1} = q^* = (w, -x, -y, -z)$.
This holds for all gates without exception: the inverse of a Hadamard, an $R_x$, a $T$
gate, or any composition thereof is computed identically. A named-gate compiler must
implement separate inversion logic for each gate type, or maintain a lookup table of
gate/inverse-gate pairs. The quaternion form makes inversion a uniform, trivially correct
operation.

\subsection{Simpler Identity Elimination}

After quaternion fusion, identity gates are quaternions near $(1, 0, 0, 0)$. Identity
detection is a single norm check: $|w - 1| < \varepsilon$ and $|x|, |y|, |z| < \varepsilon$,
for a suitable numerical tolerance $\varepsilon$. This condition is the same for every gate
type. The compiler does not need to know whether the identity arose from $H \cdot H$,
$R_z(\theta) \cdot R_z(-\theta)$, $S \cdot S^\dagger$, or any other pair.

In a named-gate compiler, identity detection requires recognizing each gate-pair pattern
separately: ($X, X$), ($H, H$), ($S, S^\dag$), ($R_z(\theta), R_z(-\theta)$), and so on.
Not all patterns can be enumerated in advance, particularly for parameterized gates where
the cancellation depends on the parameter values. The quaternion form subsumes all of these
into a single numerical test, applied uniformly to all fused quaternions. The completeness
of the identity elimination pass is therefore a consequence of the representation, not of
the compiler developer's diligence in listing all cases.

\subsection{Reduced Rewrite-Rule Surface Area}

The foregoing advantages share a common root: the quaternion IR dramatically reduces the
surface area of the optimizer's rule set. A name-based optimizer must have a rule for each
semantic relationship that the optimizer needs to exploit: fusion rules, inversion rules,
identity patterns, commutativity conditions. These rules must be stated for each gate type
and each pair of interacting gate types. The rule set grows roughly quadratically with
the number of gate types supported.

The quaternion optimizer has exactly one rule for fusion, one rule for inversion, and one
rule for identity detection. Additional optimization passes (commutation-aware ordering,
peephole optimization over longer sequences, hardware-native shortest-path decomposition)
can be stated in terms of quaternion arithmetic without reference to gate names. The
reduction in rule-set complexity is not incidental; it is a consequence of operating at
the correct level of abstraction. The quaternion IR exposes the structure that the
optimizer needs to reason about, and suppresses the syntactic variation that the optimizer
does not need to reason about.

\subsection{Clearer Separation Between Optimization and Lowering}

A well-designed compiler IR should enforce a clean boundary between the optimization
phase and the code-generation (lowering) phase. In a named-gate pipeline, this boundary
is blurred: optimization passes are aware of gate names, backend adapters may need to
re-optimize after translation, and the line between ``optimizing'' and ``translating''
is not structurally enforced.

The u1q IR enforces this boundary by construction. The optimization phase operates
entirely on quaternions and produces a quaternion-valued circuit. The lowering phase
takes that quaternion-valued circuit and produces backend-native gates, with no
optimization logic involved. Neither phase has any dependency on the other's vocabulary.
This separation makes each phase easier to implement, test, and verify independently.
It also makes the overall compiler more modular: a new backend can be added by writing
a new lowering adapter without touching any optimization pass, and a new optimization
pass can be added without touching any backend adapter.

\subsection{Summary: A Better Optimization Boundary}

Each of the advantages described in this section follows from a single design decision:
placing the optimization boundary at the level of SU(2) geometry rather than at the level
of gate names. The quaternion IR is not simply another encoding of the same information.
It is a representation that makes the relevant mathematical structure explicit and
suppresses the irrelevant syntactic structure. For the specific problem of single-qubit
gate optimization, this makes the quaternion IR a better optimization boundary than
any name-based alternative: more complete in the equivalences it captures, more uniform
in the rules it supports, more modular in the compiler architecture it enables, and
more robust to the syntactic variation introduced by a diverse and evolving hardware
ecosystem.

%% ============================================================
\section{Compiler Pipeline and Realization}
%% ============================================================

\begin{figure*}[t]
\centering
\begin{tikzpicture}[
  box/.style={rectangle, draw, rounded corners=2pt,
    minimum width=2.6cm, minimum height=0.75cm,
    align=center, font=\small},
  stdbox/.style={box, fill=gray!15},
  qbox/.style={box, fill=blue!10},
  arr/.style={-{Stealth[scale=1.1]}, thick},
  lbl/.style={font=\small\bfseries, anchor=east}
]

% Top lane: standard named-gate pipeline
\node[lbl] at (-0.3, 0.9) {Standard:};
\node[stdbox] (s1) at ( 1.4, 0.9) {Named\\Gates};
\node[stdbox] (s2) at ( 4.2, 0.9) {Gate Alias\\Table};
\node[stdbox] (s3) at ( 7.0, 0.9) {Syntax\\Rewriting};
\node[stdbox] (s4) at ( 9.8, 0.9) {Backend-Specific\\Handling};
\node[stdbox] (s5) at (12.6, 0.9) {Output};

\draw[arr] (s1) -- (s2);
\draw[arr] (s2) -- (s3);
\draw[arr] (s3) -- (s4);
\draw[arr] (s4) -- (s5);

% Bottom lane: quaternion IR pipeline
\node[lbl] at (-0.3, -0.3) {Quaternion IR:};
\node[qbox] (q1) at ( 1.4, -0.3) {Named\\Gates};
\node[qbox] (q2) at ( 4.2, -0.3) {Quaternion\\Normalization};
\node[qbox] (q3) at ( 7.0, -0.3) {Semantic\\Optimization};
\node[qbox] (q4) at ( 9.8, -0.3) {Backend\\Lowering};
\node[qbox] (q5) at (12.6, -0.3) {Output};

\draw[arr] (q1) -- (q2);
\draw[arr] (q2) -- (q3);
\draw[arr] (q3) -- (q4);
\draw[arr] (q4) -- (q5);

\end{tikzpicture}
\caption{Comparison between syntax-level gate optimization and quaternion-based semantic
normalization. The quaternion IR collapses backend-specific gate aliases into a single
optimization boundary before lowering.}
\label{fig:pipeline}
\end{figure*}

The compiler pipeline translates the quaternion IR principle into a concrete sequence of
passes, providing evidence that the approach is practically realizable. The pipeline
passes are designed so that each one operates on a well-defined IR form and produces
a well-defined IR form, with no pass requiring knowledge of another pass's internal logic.

Gate resolution is the first pass. It maps named gates to their exact SU(2)
correspondences using the quaternion math primitives described in Section~III. For
parameterized gates such as $R_x(\theta)$, this involves evaluating the trigonometric
expressions for $w$ and the relevant component. For multi-angle gates such as
$U(\theta, \phi, \lambda)$, this involves computing the ZYZ Euler decomposition and
extracting quaternion components from the resulting SU(2) matrix. The output of gate
resolution is a mixed IR containing u1q nodes for single-qubit gates and unchanged nodes
for two-qubit gates.

The \texttt{to\_u1q\_pass} completes the normalization step by verifying that all
single-qubit nodes satisfy the u1q invariants in Table~\ref{tab:invariants}. This
includes enforcing $w \geq 0$ by negating any quaternion with $w < 0$, and checking
that the unit-norm constraint is satisfied to within the numerical tolerance. After this
pass, the IR is fully in canonical u1q form and ready for optimization.

Quaternion fusion collapses consecutive u1q gates on the same qubit via
\eqref{eq:fusion}, re-canonicalizing after each multiplication to maintain the $w \geq 0$
invariant. Fusion traverses each qubit's single-qubit segment --- the maximal sequence
of single-qubit operations between consecutive two-qubit gate boundaries --- and reduces
it to a single quaternion. The fused quaternion represents the exact composition of all
single-qubit operations in that segment. If the segment contains $n$ gates, the fused
result is computed in $n-1$ quaternion multiplications, each a constant-time operation.

Identity elimination removes any fused gate satisfying $q \approx (1, 0, 0, 0)$. The
tolerance is chosen conservatively to avoid false positives from accumulated floating-point
error, while still eliminating gates that are physically redundant. Geodesic canonicalization
enforces $w \geq 0$ globally, selecting the shorter arc on $S^3$ and improving numeric
stability for backend decomposition.

Backend lowering is performed once per adapter. Each backend adapter (Qiskit, Amazon
Braket, PennyLane) receives the optimized u1q IR and translates to its native rotation
primitive using a fixed decomposition. No backend re-derives gate semantics; each adapter
implements a common interface contract that accepts a u1q quaternion and returns a sequence
of backend-native gates. Because all optimization has already been applied at the u1q level,
the backend adapter's output is already as short as the optimizer can make it.

The pipeline is realized as a Python library and REST API, providing a reproducible
optimization service for evaluation purposes. The same input circuit yields the same
optimized output on every invocation, because the pipeline is determined entirely by
quaternion arithmetic and contains no non-deterministic passes. This reproducibility is
itself a consequence of operating on a mathematically well-defined canonical form.

%% ============================================================
\section{Evaluation}
%% ============================================================

We evaluate the optimization pipeline on a set of circuits designed to exercise the
implemented passes. The evaluation serves two purposes: it validates that the
implementation is correct (optimized circuits are semantically equivalent to original
circuits), and it illustrates the type of optimization the quaternion IR enables. The
evaluation is not intended to establish performance claims relative to fully optimized
industrial compilers; the scope of the current pipeline is more focused, and a broader
comparative evaluation is planned for future work.

Rotation-heavy circuits test quaternion fusion: they consist primarily of single-qubit
rotations that can be composed into a smaller number of equivalent rotations. Standard-gate
circuits test identity elimination: they consist of gate pairs that compose to the identity
(e.g., $H \cdot H$) and should be removed entirely. All optimized circuits pass unitary
equivalence verification: $\|U_{\text{original}} - U_{\text{optimized}}\|_F < 10^{-9}$,
where the Frobenius norm test is applied to the full circuit unitary computed by direct
matrix product.

Two-qubit gate counts are unchanged in all cases, confirming that reported reductions
are caused by the single-qubit passes and not by incidental effects of backend translation.
Table~\ref{tab:results} summarizes results for three benchmark circuits where complete data
are available.

\begin{table}[htbp]
\caption{Benchmark Results Summary}
\begin{center}
\begin{tabular}{p{2.2cm}ccccccc}
\toprule
\textbf{Circuit} & \textbf{Qb} &
\textbf{G\textsubscript{base}} & \textbf{G\textsubscript{opt}} & \textbf{Red.} &
\textbf{D\textsubscript{base}} & \textbf{D\textsubscript{opt}} & \textbf{Eq.} \\
\midrule
H${\cdot}$H chain ($n{=}4$) & 1 & 4 & 0 & 100\% & 4 & 0 & \checkmark \\
Rx${\cdot}$Rx seq.           & 1 & 2 & 1 &  50\% & 2 & 1 & \checkmark \\
Bell prep                    & 2 & 3 & 3 &   0\% & 2 & 2 & \checkmark \\
\bottomrule
\multicolumn{8}{l}{\footnotesize Qb = qubits; G = gate count; D = depth; Eq. = semantic equivalence.}
\end{tabular}
\label{tab:results}
\end{center}
\end{table}

The H$\cdot$H chain result demonstrates identity elimination: four Hadamard gates in
sequence collapse to zero gates after quaternion fusion followed by identity detection.
The $R_x \cdot R_x$ result demonstrates angle fusion: two $x$-axis rotations are combined
into one, halving the single-qubit gate count. The Bell preparation result is expected
behavior: the circuit contains no redundant single-qubit sequences, so the single-qubit
passes produce no reduction. This last case confirms that the pipeline does not alter
circuits where no optimization opportunity exists, a necessary correctness check.

The key claim of this paper is structural and compiler-theoretic rather than primarily
quantitative. The claim is that the quaternion IR provides a better optimization boundary
than named-gate representations, in the sense that it is more semantically complete,
more uniform in its rule set, and more modular in the compiler architecture it implies.
These properties are not established by benchmark numbers alone; they are consequences
of the algebraic structure of SU(2) and the isomorphism with unit quaternions. The
benchmark results provide evidence that the implementation correctly realizes the IR
concept and that the optimization passes function as intended. Broader quantitative
evaluation --- covering variational circuits, Clifford circuits, and application-scale
benchmarks --- is ongoing and will form the basis of a subsequent evaluation study.

%% ============================================================
\section{Related Work}
%% ============================================================

\subsection{Mainstream Transpiler Pipelines}

Qiskit's transpiler \cite{b1} applies optimization at multiple levels (O0--O3), including
routing, basis translation, and peephole rewriting. PennyLane \cite{b2} applies
device-specific compilation and supports differentiable optimization. Amazon Braket \cite{b3}
provides device-level compilation through provider SDKs. These pipelines share a common
structure: optimization applies pattern-matching and rewrite rules over named-gate syntax.
Each optimization pass in these pipelines must be aware of the gate names present in the
circuit, and each new backend or gate alias requires updating the relevant passes. The
quaternion IR approach differs structurally by establishing a geometric normalization
boundary \textit{before} any optimization pass runs. Rather than extending the optimization
passes to handle new gate names, the quaternion approach requires only updating the
gate-resolution step that maps new names to their quaternion form. This architectural
difference becomes more significant as the number of supported backends and frameworks
grows.

\subsection{IR-Centric Compiler Approaches}

The MLIR \cite{b4} ecosystem and its quantum extensions demonstrate the value of multi-level
IRs for quantum compilation. Catalyst \cite{b5} introduces multiple abstraction levels and
progressive lowering for differentiable quantum programs. These frameworks establish the
general principle that a well-chosen IR boundary improves compiler modularity and
correctness. The quaternion IR is a specific canonical form at the single-qubit level,
complementary to multi-level IR frameworks rather than competing with them. Where MLIR-style
approaches define abstraction hierarchies across multiple levels of the compilation stack,
the u1q IR is a targeted geometric normalization at a single level: the boundary between
named-gate input and semantic optimization. The two design principles can coexist: a
multi-level quantum compiler could incorporate u1q as its single-qubit normalization
layer within a broader multi-dialect IR architecture.

\subsection{Gate Rewriting and Peephole Optimization}

Template-based optimization \cite{b8} identifies and replaces fixed gate patterns,
typically using pre-computed equivalence tables. This approach is complementary to
quaternion fusion for short fixed sequences, but does not generalize to arbitrary
parameterized rotations. Clifford circuit optimization \cite{b6,b7} applies ZX-calculus
rewrite rules to reduce Clifford gate counts. PyZX \cite{b9} applies graph-based rewriting
using the ZX-calculus. These approaches operate at a higher level of abstraction than
the quaternion IR: they target specific gate subsets (Clifford gates) or specific
structural properties (graph-rewrite completeness), and they are more expressive than
the quaternion IR for those targets. The quaternion IR targets the complementary domain:
the continuous, parameterized single-qubit gate segment, which does not reduce well to
discrete pattern matching. The two approaches address different parts of the optimization
problem and could be composed in a pipeline that applies quaternion fusion first and
ZX-based Clifford simplification second.

\subsection{Quaternion Representations in Quantum Computing}

Quaternions have appeared in quantum computing contexts primarily for state visualization
(representing Bloch sphere rotations) and for robotics-inspired gate decomposition
algorithms. Their use as a \textit{compiler IR} --- as an algebraic normalization layer
through which all optimization passes are defined --- is, to our knowledge, not
established in prior work. The nearest prior use is in parameterized circuit optimization
where trigonometric identities are applied to simplify rotation angles; this shares the
goal of reasoning about rotations mathematically, but does not establish a canonical
IR form or use quaternion multiplication as the core fusion mechanism. The contribution
here is not the use of quaternions per se, but the engineering decision to make the
quaternion form the primary representation at the optimization boundary, and to build
a complete optimization pipeline --- with defined invariants, pass structure, and
backend interface contract --- on top of it.

%% ============================================================
\section{Limitations and Conclusion}
%% ============================================================

\subsection{Limitations}

The quaternion IR and all implemented passes operate exclusively on single-qubit gate
segments. Circuits where two-qubit gates dominate the total gate count will see minimal
improvement from the current pipeline. The analysis of two-qubit gate reduction
requires different mathematical machinery (the KAK decomposition or Weyl chamber
representation), and the quaternion IR as defined does not address this domain. This is
a deliberate scope limitation, not an architectural incompatibility: the u1q IR could
in principle be extended or composed with a two-qubit optimization layer, but that
extension is not part of the current work.

Decomposing u1q to backend-native gate sets introduces additional gates in some cases.
A ZYZ decomposition requires up to three basis gates on some backends, which can partially
offset gains from fusion and identity elimination in circuits with few redundant
single-qubit sequences. For circuits where the optimization reduction is modest (one or
two gates saved), the decomposition overhead may be comparable to or exceed the savings.
Shortest-path decomposition strategies that minimize the number of native gates produced
for a given quaternion are a known problem with established solutions \cite{b8}, but their
integration into the current pipeline is ongoing.

The current evaluation covers a small set of benchmark circuits chosen to exercise the
implemented passes. This is sufficient to validate correctness but not to draw conclusions
about average-case performance across realistic application circuits. Circuits with complex
entanglement structure, many small-angle rotations, or specific hardware constraints may
behave differently from the benchmarks reported. A more comprehensive evaluation, including
comparison against fully optimized Qiskit transpiler pipelines, is deferred to future work
and requires a broader benchmark suite.

Several planned extensions --- commutation-aware pass ordering, peephole optimization
over longer gate sequences, and hardware-native shortest-path decomposition --- are not
yet implemented. These extensions would strengthen the optimizer's ability to find
reduction opportunities in more complex circuits, but their absence does not affect the
validity of the core claim: that the quaternion IR provides a better optimization boundary
than named-gate representations for the passes that are implemented.

\subsection{Conclusion}

We have argued that a quaternion-based intermediate representation is a more principled
optimization boundary for single-qubit quantum circuits than standard named-gate pipelines.
The quaternion IR normalizes all single-qubit gates into one canonical SU(2) form before
optimization, collapsing gate aliases, unifying fusion and inversion rules, and enabling
backend-agnostic optimization. These advantages are structural: they follow from the
algebraic isomorphism between unit quaternions and SU(2), not from heuristics or
backend-specific engineering. Any optimization pass that can be stated in terms of quaternion
arithmetic automatically applies to the complete space of single-qubit gates, in all
frameworks and parameter conventions, without modification.

The current implementation demonstrates that this approach is practically realizable:
the canonicalization invariants are well-defined, the pass sequence is straightforward
to implement, and the results are semantically verifiable. The current gains are strongest
for single-qubit-dominated circuits. Extending the optimization boundary to two-qubit
operations and integrating with hardware-native compilation stages remain open problems
for future work.

More broadly, the quaternion IR illustrates a general design principle: compiler quality
for quantum circuits is improved by locating optimization boundaries at the level of
algebraic structure rather than at the level of syntactic convention. The unit quaternion
representation of SU(2) is a natural, compact, and computationally convenient realization
of this principle for single-qubit gates. It provides a stable and principled semantic
layer on which a more complete quantum compiler optimization infrastructure can be built.

%% ============================================================
\begin{thebibliography}{00}

\bibitem{b1}
Qiskit contributors, ``Qiskit: An open-source framework for quantum computing,'' 2024.
[Online]. Available: https://github.com/Qiskit/qiskit

\bibitem{b2}
V. Bergholm et al., ``PennyLane: Automatic differentiation of hybrid quantum-classical
computations,'' arXiv:1811.04968, 2018.

\bibitem{b3}
Amazon Web Services, ``Amazon Braket,'' 2020. [Online]. Available:
https://aws.amazon.com/braket/

\bibitem{b4}
C. Lattner et al., ``MLIR: Scaling compiler infrastructure for domain specific computation,''
in \textit{Proc. IEEE/ACM CGO}, 2021.

\bibitem{b5}
V. Bergholm et al., ``Catalyst: A Python-first compiler for quantum programs,'' PennyLane
documentation, 2024.

\bibitem{b6}
S. Bravyi, S. Hu, D. Maslov, and R. Shaydulin, ``Clifford circuit optimization with templates
and symbolic Pauli gates,'' \textit{Quantum}, vol. 5, p. 580, 2021.

\bibitem{b7}
R. Duncan, A. Kissinger, S. Perdrix, and J. van de Wetering, ``Graph-theoretic simplification
of quantum circuits with the ZX-calculus,'' \textit{Quantum}, vol. 4, p. 279, 2020.

\bibitem{b8}
D. Maslov, G. W. Dueck, D. M. Miller, and C. Negrevergne, ``Quantum circuit simplification
and level compaction,'' \textit{IEEE Trans. Comput.-Aided Design Integr. Circuits Syst.},
vol. 27, no. 3, 2008.

\bibitem{b9}
A. Kissinger and J. van de Wetering, ``PyZX: Large scale automated diagrammatic reasoning,''
in \textit{Proc. QPL 2019}, EPTCS 318, 2020.

\end{thebibliography}

\end{document}
